---
title: "Session 4: Word frequencies and visualization"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
bibliography: textanalysis.bib
---
```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

# Tidy corpus form

We've talked about tidy format -- each variable a column, each observation a row, each type of observational unit a table. We've taken speeches at the Nobel prize and pages of oil company Sustainability Reports as our observations and can use this to do some transformations -- do total word counts or specific word frequencies. But once we start really doing analysis at the level of the word, tidy rules would seem to dictate that we treat not individual texts but individual _words_ as observations. Thus tidytext format is a table with one word per row. 

It might help if we are more specific: we can use two signifiers for words - words and tokens. Words are unique, well, words as we conceive of them. Tokens are each individual instances of a word. So a sentence: "the brown fox jumped over the brown log" we will say has 8 tokens (number of total words) but 6 words (number of unique words, as "the" and "brown" appear twice).^[We could also start talking about tokens and words being not just things we'd recognize as words but things like :) and so on.] So tidy data format is *one token per row*.^[We will also later talk about n-grams where we might take tokens to be 2-grams, sentences or even longer pieces of text.] 

We convert character strings that we have been working with into tidy text with the ``unnest_rows()`` function in the ``tidytext package``, which splits texts up by tokens (a process appropriately called "tokenization").

```{r, message=FALSE}
# Lets use the first speach in our Nobel corpus as an example
library(tidytext)
library(tidyverse)
nobel <- read_rds("data/nobel_cleaned.Rds")
ex <- nobel[1,]
ex %>%
  unnest_tokens(output = words, input = AwardSpeech)

```

We now end up with a dataframe where each row is an observation with words, year, and Nobel laureate as variables. Note also that ``tidytext`` does cleaning - whitespace stripping, to_lower and so on. We didn't notice here as we're using our dataframe that we have already cleaned. The first thing we might notice is the number of rows is the total word count. 

# Stop Words

Another way in which texts are frequently pre-processed for analysis and we can now look at is to remove so-called stop words. Stop words are words with grammatical rather than syntactic function - they make things we say grammatical but don't add (much) meaning. Examples include "a", "the", "of" and so on. What counts as a stopword might vary on our corpus! And what sort of analysis we're trying to do.^[Interestingly, one of the most famous cases of computational text analysis in the social science -- Mosteller and Wallace's attribution of anonymously published Federalist papers -- analyzed stopwords and threw out everything else (@mosteller_inference_1963).] All major text analysis packages have means for removing stop words. In ''tidytext'' we have a list of stopwords called ``stop_words`` from a couple of different corpora (``?stop_words`` for more info and links).

```{r}
stop_words
nobel_tidy <- nobel %>%
  unnest_tokens(output = words, input = AwardSpeech) %>%
  anti_join(stop_words, by = c("words" = "word"))  # by= specifies which columns to use, had they been named the same thing we could have omitted it
```

``stop_words``, we'll recall, is a tibble, so we can easily make our own tibble of stopwords and add it. Say (for purposes of demonstration) I think that the Peace Prize award ceremony is, of course, going to talk about peace and war so we want to weed these words out of our frequency counts.

```{r}
my_words <- c("peace", "war")
custom_stop_words <- tibble(word = my_words, lexicon = "my_customization")
stop_words_custom <- rbind(stop_words, custom_stop_words)
tail(stop_words_custom) # view the end of the tibble, look like our words were added correctly
```

Now we can apply the ``stop_words_custom`` just like we did ``stop_words``. We won't actually do this because we probably want these words in the corpus!

# Stemming

If we are looking at word frequencies, will we want to count "represent" and "represented" as the same word? Or "war" and "wars"? Very likely. If so we can transform our text via a process called stemming, cutting down words to their stems so that different forms of these word are recognized as being the same thing. Something like this matters in English but might really matter in an more highly inflected language like Russian. 

We have a couple stemmers to choose from in R, one and the best known is the Porter Stemming [algorithm](https://tartarus.org/martin/PorterStemmer/). Another is [hunspell](https://docs.ropensci.org/hunspell/articles/intro.html), based on the popular open source and multilingual spell checker. We'll use Porter here.

To see it in action we'll first test it against a short test set of words and then apply to our whole corpus.

```{r, message=FALSE}
library(SnowballC)
words_to_stem <- c("going", "represented", "wars", "similarity", "books")
SnowballC::wordStem(words_to_stem)
```

Stemming in action. So now applied to the entire document:

```{r}
(nobel_tidy_stemmed <- nobel_tidy %>%
  mutate(word_stem = wordStem(words)))
write_rds(nobel_tidy_stemmed, "data/nobel_stemmed.Rds")
```

# Top word frequencies^[This section and the next lean heavily on chapters 1 and 3 in @silge2017text]

Now we can find the most frequently appearing words in the corpus.

```{r}
nobel_tidy %>%
  count(words, sort=TRUE)
```

Remember we got this nice and informative result only because we already removed the stopwords. OTherwise we would have been swamped with "the", "and", and so on.

Our knowledge of how to subset tibbles will now come in pretty handy. If we want to get the most frequent words, say, before 1945 we easily do this.

```{r}
nobel_tidy %>%
  filter(Year < 1945) %>%
  count(words, sort=TRUE)
```

And we could then compare it to post-1945 word frequency and we'd see pre-WWII prize speeches were more concentrated on Europe, the League, and nations while after that we get more talk of world, people, human, committee. That makes sense.

Let's graph this.

```{r}
nobel_tidy %>%
  count(words, sort=TRUE) %>%
  top_n(15) %>%                     # selecting to show only top 15 words
  mutate(words = reorder(words,desc(n))) %>%  # this will ensure that the highest frequency words appear to the left
  ggplot(aes(words, n)) +
    geom_col()
```

And with just a little bit more code we can view pre-1945 and post-1945 top frequency words at the same time.

```{r}
nobel_tidy %>%
  mutate(Period = ifelse(Year <= 1945, "Pre-WWII", "Post-WWII")) %>%   # creating columns with label "Pre-WWII" and "Post-WWII"
  mutate(Period = factor(Period, levels = c("Pre-WWII", "Post-WWII"))) %>%
  group_by(Period) %>%                                                # grouping by this column label so frequencies will be                                                                              calculated within group
  count(words, sort=TRUE) %>%
  mutate(proportion = n / sum(n) * 1000) %>%                     # perhaps we'd like word frequency per 1000 words rather than raw                                                                      counts?
  slice_max(order_by=proportion, n = 15) %>%                     # selecting to show only top 15 words within each group
  ggplot(aes(reorder_within(x = words, by = proportion, within = Period), proportion, fill = Period)) +    # reordering is a bit tricky, see                                                                                                     ?reorder_within()
    geom_col() +
    scale_x_reordered() +
    coord_flip() +
    facet_wrap(~Period, ncol = 2, scales = "free") +
  xlab("Word")
```

## Word Clouds

One common visualization of word frequency is word clouds. To do this we use the package wordcloud which will work very nicely with our tidily organized data. Wordcloud2 gives color and more fancy options that you can also play with.

```{r}
library(wordcloud)
library(wordcloud2)
nobel_tidy %>%
  count(words, sort=TRUE) %>%
  with(wordcloud(words, n, max.words = 100))

dat <- nobel_tidy %>%
  count(words, sort=TRUE) %>%
  mutate(word = words) %>%
  mutate(freq = n) %>%
  select(word, freq) %>%
  top_n(200)
wordcloud2(dat, size = 2)
```

It's also possible to do word clouds that compare two documents. To do this we'll need to step outside the tidyverse and organize our data in other formats so we save this for Session 5. 

# TF-IDF 

As we've seen, there are multiple ways to calculate frequency -- we can take raw counts, or term frequency ($tf$). Proportions, term frequency divided by total token count in a given document, are another means. Certainly proportions make it easier to compare across corpora of different size. The problem with this is that they tend to get flooded with stopwords. One means of dealing with this is to remove the stopwords as we have done, but another is to attempt to downweigh words that appear often everywhere and upweigh those that are more unusual. Inverse document frequency ($idf$) is a weighting system to do this -- it equals the total number of documents in the corpus divided by the number of documents in the corpus that contain the given word. The greater the number of documents in the corpus in which the word does _not_ appear (suggesting words that are unique to certain documents rather than widespread across the corpus as a whole) the smaller the denominator and, thus, the greater the ratio.

TF-IDF is is term frequency times inverse document frequency. Both are often logged. In symbols,

\begin{equation}
   \text{TF}_{t,d} =
 \begin{cases}
   1 + \text{log}_{10} \: \text{count(t,d)}, & \text{if count(t,d)} > 0 \\
   0,&\text{otherwise.}\\
 \end{cases}
 \end{equation}

 \begin{equation}
   \text{IDF}_{t} = \text{log}_{10} \: \bigg(\frac{N}{\text{df}_t}\bigg).
 \end{equation}

where t is the given term, d is a given document, df$_t$ is the number of documents in the corpus containing term t, and N is the total number of documents in the corpus. Long story short, tf-idf attempts to weight words by both frequency in an individual document and their unusualness over a corpus of documents. Every word in every document will have its own tf-idf (term frequency will vary across documents while inverse document frequency is the same across the corpus).

Let's see how we do this in R. First we'll compute document frequency. In order to simplify results, lets use the same subsetting of our data into pre-1945 and post-1945 -- this means we're treating pre-WWII speeches as one single document and likewise post-1945 speeches. `tidytext` makes it pretty easy -- simply unnest tokens and then count the tokens. Note that ``count`` enable counting within groups, which we passed to ``count`` telling it to do the counts within the groups denoted in column ``Period``. This produces the same result as passing ``group_by(Period)`` in the previous line and eliminating ``Period`` from the ``count()`` call.

```{r, message = FALSE}
nobel <- read_rds("data/nobel_cleaned.Rds") %>%
  mutate(Period = ifelse(Year <= 1945, "Pre-WWII", "Post-WWII"))
nobel_words <- nobel %>%
  unnest_tokens(words, AwardSpeech) %>%
  count(words, Period, sort = TRUE)
```

We can then use ``bind_tf_idf()`` from tidytext. (Tidytext implements tf-idf using proportional, but not logged, tf -- we'll see some of these other versions in other packages). The function takes a first argument (other than the tidy dataframe) that is the word, a second that is the document, and third a column containing document-term counts). 

```{r}
tf_idf <- nobel_words %>%
  bind_tf_idf(words, Period, n)
tf_idf
```

Here we see that tf-idf has zeroed out these extremely common and not very interesting terms, precisely what we'd hope an indicator like this would do. Lets see the highest tf-idf scores.

```{r}
tf_idf %>% arrange(desc(tf_idf)) 

tf_idf %>%
  mutate(Period = factor(Period, levels = c("Pre-WWII", "Post-WWII"))) %>%
  group_by(Period) %>%
  slice_max(tf_idf, n = 15) %>%
  ungroup() %>%
  ggplot(aes(tf_idf, fct_reorder(words, tf_idf), fill = Period)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Period, ncol = 2, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```

Many of these are names, logically enough as specific names of laureates appear mostly only when they are being awarded a prize thus increasing their idf and upweighing their tf-idf. To make this more useful we'd want to go through and week out (remove, just like stopwords) names. But still we see nuclear coming to the fore in the post-war era, reasonably enough! Things like "global", "poverty" stand out post-war while pre-war we see "reparations", "commercial", more Europe specific vocabulary. A different conversation pre- and post-war.

# References
